---
title: "Lecture 16"
subtitle: "Correlation and Linear Regression"
author: "Psych 10 C"
institute: "University of California, Irvine"
date: "05/06/2022"
output:
  xaringan::moon_reader:
    self_contained: true
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-extra, echo = FALSE}
xaringanExtra::use_tile_view()

xaringanExtra::use_fit_screen()

xaringanExtra::use_editable(expires = 1)

xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         #<<
  mute_unhighlighted_code = TRUE  #<<
)

htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

```{r load-tidy, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)

knitr::opts_chunk$set(fig.width = 6, message = FALSE, 
                      warning = FALSE, comment = "", 
                      cache = F)
```

## Correlation

- Up to this point we have looked at problems with a single dependent variable, 
however, we could have more than one.

--

- When we have 2 dependent variables we could ask ourselves, how are those two 
variables related?

--

- The correlation between two variables is a measure of the association between 
them.

--

- Formally, it's a measure of how the two variables change together. For 
example, what happens to the values of one variable as the second one increases?

--

- The correlation coefficient measures this degree of association and we denote
this value with $R$.

---

## Correlation

- The correlation coefficient can take positive or negative values or it can 
be 0.

--

- However, it will always be bounded between -1 and 1, for example:

.pull-left[
```{r cor-1,echo=FALSE, fig.align='center'}
iq <- rnorm(50, 100, 10)
a <- 15 - 3/20 * 100
b <- 3/20
plot(iq, a + b * iq, axes = F, ann = F, pch = 16,
     col = "#0D95D0", cex = 1.3)
box(bty = "l")
axis(side = 1)
axis(side = 2, las = 2)
mtext(text = "IQ", side = 1, line = 2.6, cex = 1.8)
mtext(text = "Depression", side = 2, line = 2.6, cex = 1.8)
mtext(text = "R = 1", cex = 2, side = 3)
```
]
.pull-right[
```{r cor-2,echo=FALSE, fig.align='center'}
a <- 15 + 3/20 * 100
b <- 3/20
plot(iq, a - b * iq, axes = F, ann = F, pch = 16,
     col = "#0D95D0", cex = 1.3)
box(bty = "l")
axis(side = 1)
axis(side = 2, las = 2)
mtext(text = "IQ", side = 1, line = 2.6, cex = 1.8)
mtext(text = "Depression", side = 2, line = 2.6, cex = 1.8)
mtext(text = "R = -1", cex = 2, side = 3)
```
]

---

## Correlation

- A correlation of 0 should look like this:

```{r cor-0, echo=FALSE,fig.align='center', fig.width=7}
x <- MASS::mvrnorm(200,c(15,100), c(1,10) * diag(2))
plot(x[,2], x[,1], axes = F, ann = F, pch = 16,
     col = "#0D95D0", cex = 1.3)
box(bty = "l")
axis(side = 1)
axis(side = 2, las = 2)
mtext(text = "IQ", side = 1, line = 2.6, cex = 1.8)
mtext(text = "Depression", side = 2, line = 2.6, cex = 1.8)
mtext(text = "R = 0", cex = 2, side = 3)
```

---

## Correlation

- Therefore, the sign of the correlation indicates whether one variable 
**increases** as the otherone **increases** (**positive R**) or if one variable 
**decreases** as the other one **increases** (**negative R**).

--

- On the other hand, the magnitud of R indicates the "strength" of the 
association, with values closer to 1 (or -1) representing a stronger association
between the variables.




