---
title: "Lecture 4"
subtitle: "Statistical Models and Notation"
author: "Psych 10 C"
institute: "University of California, Irvine"
date: "04/06/2022"
output:
  xaringan::moon_reader:
    self_contained: true
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-extra, echo = FALSE}
xaringanExtra::use_tile_view()

xaringanExtra::use_fit_screen()

xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         #<<
  mute_unhighlighted_code = TRUE  #<<
)

htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

```{r load-tidy, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)

library(flipbookr)

knitr::opts_chunk$set(fig.width = 6, message = FALSE, 
                      warning = FALSE, comment = "", 
                      cache = F)

link <- "https://raw.githubusercontent.com/ManuelVU/psych-10c-data/main/example.csv"

memory <- read_csv(file = link)

memory <- memory %>% 
  mutate("test_id" = ifelse(test = time_test == 300, 
                            yes = "test_1", no = "test_2")) %>%
  filter(test_id == "test_1")
```

## Objective in research

- One of our main objectives in research is to contrast our believes about the
world with the outcomes of experiments.

--

- We do so by starting with some "verbal" statement or belief about the world 
which we then formalize using a statistical model.

--

- Statistical models will allow us to make predictions about future 
observations. In the case of an experiment, they will allow us to make 
predictions about the outcomes.

--

- The next step is to evaluate those predictions by comparing them with the 
outcomes (data) of the experiment.

--

- Finally, we would like to go back and interpret the results of our evaluations
with respect to our original believes or statements about the world.

---
 
## Statistical models

- Statistical models are abstract representations of the world. 

--

- They are a way in which we can formalize our believes about probabilistic 
events.

--

- For example, if we have an experiment where we throw a coin and have two 
competing ideas about the coin:

--

  - The coin is **fair**.
  
--

  - The coin is **not fair**.

--

- We can formalize these two believes using a statistical model.

--

  - The coin is fair: $P(\{heads\})\ =\ P(\{tails\})\ =\ 0.5$
  
--

  - The coin is not fair: $P(\{heads\})\ \neq\ P(\{tails\})$
  
--

- We moved from two verbal statements about our believes regarding the coin to
two formal statements about the probability of "heads".
  
---

## Statistical Models

- Statistical models are the formal representation of our believes or hypothesis
about the outcomes of an experiment. 

--

- Given that we assume that the outcomes are probabilistic, our models will have 
a probabilistic component associated with them. 

--

- Given the nature of our observations it will be almost impossible for us to 
tell if a model is TRUE or FALSE. However, we can compare how useful they are on 
a given situation. 

--

- Statistical models will allow us to make predictions about our observations, 
which we will then use to compare how useful they are.

--

- However, before we continue it will be useful to introduce some notation!

--

- This will provide us with a way to express our models in a formal and 
standard way.

---

class: inverse, middle, center

# Notation

---

## Example:

- To introduce notation we will start with a problem.

--

- **Problem:** We want to know if people that smoke have lower lung capacity in 
comparison with people that do not smoke.

--

- We have a variable that we are interested in, which is lung capacity as 
measured by some standard test.

--

- We also have a variable that indicates if a given participant smokes or not.

--

- We call the first one a **dependent** variable, because we want to see how it 
"depends" on the values of another.

--

- We call the smoker indicator variable an **independent** variable. We are 
interested in how our independent variable affects the values of our dependent 
variable.

--

- In other words, we want to know if lung capacity is a function of smoking 
status.

---

## Example: Smoking

- We collect data from 8 participants, 4 smokers and 4 non smokers.

--

```{r, echo = FALSE}
smokers <- tibble("id" = as.character(seq(1,8)),
  "lung_capacity" = round(x = rnorm(n = 8, 
                                    mean = rep(x = c(77, 70), each = 4), 
                                    sd = rep(x = c(2, 3), each = 4)), 
                          digits = 1),
  "smoke_status" = rep(c("non_smoker", "smoker"), each = 4))
DT::datatable(data = smokers,
  fillContainer = FALSE, options = list(pageLength = 4)
)
```

--

- We will denote values of our dependent variables using $y$ for example, the 
first observation of our first group (non-smokers) is denoted as $y_{11}$ while 
the fourth observation of the same group is denoted $y_{41}$

---

- In general we say that the *i-th* observation of the *j-th* group is denoted 
as $y_{ij}$. Note that the letters $i$ and $j$ are a way to denote a general 
observation, if we want to look at a particular one we can write $y_{21}=$ 
`r smokers$lung_capacity[2]`.





