---
title: "Lecture 12"
subtitle: "Models for Factorial Designs"
author: "Psych 10 C"
institute: "University of California, Irvine"
date: "04/25/2022"
output:
  xaringan::moon_reader:
    self_contained: true
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-extra, echo = FALSE}
xaringanExtra::use_tile_view()

xaringanExtra::use_fit_screen()

xaringanExtra::use_editable(expires = 1)

xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         #<<
  mute_unhighlighted_code = TRUE  #<<
)

htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

```{r load-tidy, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)

knitr::opts_chunk$set(fig.width = 6, message = FALSE, 
                      warning = FALSE, comment = "", 
                      cache = F)

link <- "https://raw.githubusercontent.com/ManuelVU/psych-10c-data/main/anxiety-w5.csv"

students <- read_csv(link) %>%
  mutate(id = as.character(id),
         anxiety = as.integer(anxiety),
         cohort = as.factor(cohort))
  
```

## Models for factorial designs

- The previous class we talked about two different models for factorial designs.

--

- The **Null** model which formalizes the assumption that the combinations of our
factors (groups) have no effect on the expectation of our dependent variable 
(observations).

--

- The Null model is expressed formally as:
$$y_{ijk}\sim\text{Normal}(\mu,\sigma_0^2)$$

- Where $i$ represents the observation number of the combination of the *j-th* 
level of factor 1 and the *k-th* level of factor 2.

--

- The second type of model we covered was the **Main effects** model. Main 
effects models assume that the expected value of our dependent variable is 
different between levels of a single factor regardless of the values of other 
factors on the experiment.

--

- The number of Main effects models that we will have depends on the number of 
independent factors that we have.

---

## Models for factorial designs

- As we saw last class, a Main effects model for factor $j$ was expressed as:
$$y_{ijk}\sim\text{Normal}(\mu+\alpha_j,\sigma_1^2)$$

--

- While the main effects model of factor $k$ is:
$$y_{ijk}\sim\text{Normal}(\mu+\beta_k,\sigma_2^2)$$

--

- We will only work with $2\times 2$ factorial designs for now, so these are the
only two main effects models that we need.

--

- Remember that we use a different effects variable $(\alpha_j$ for factor $j$ 
and $\beta_k$ for factor $k$) because we will use those variables for another 
model.

--

- Today, we will introduce the remaining 2 models that we use in a factorial
design and work on an example using data from our anxiety example with different
cohorts of students that took either a statistics course during the year or 
another class.










