---
title: "Lecture 1"
subtitle: "Syllabus, Review R"
author: "Psych 10 C"
institute: "University of California, Irvine"
date: "03/28/2022"
output:
  xaringan::moon_reader:
    self_contained: true
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-extra, echo = FALSE}
xaringanExtra::use_xaringan_extra(c("title_view", "fit_screen"))

xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         #<<
  mute_unhighlighted_code = TRUE  #<<
)

htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```


class: middle, center

# Welcome to Psych 10 C

---

## Psych 10 C

--

- This is the third course in the series.

--

- Instructor: Manuel Villarreal <p style="color:purple;">jesusmv1@uci.edu</p> 

--

- TA's: 
  - Adriana <p style="color:purple;">achavezd@uci.edu</p>
  - Jeff <p style="color:purple;">jcoon1@uci.edu</p>

---

## Structure of the course

- Goal: learn how and when to use linear models to answer questions about data.

--

- Evaluation:
  - 5 homework assignments
  - 1 final project
  
--

- Both require the use of R code, however, this is not a coding class so you 
will be able to find what you need on the slides.

--

- We will provide homework templates that you will need to fill in (except for
the final which will be an empty template).

---

## Grading:

- Each homework will be worth 10 points while the final project is worth 50, for
a total of 100 points.

--

- Grades will be assigned as follows:

| Grade| Interval     | Grade | Interval     | Grade | Interval   |
|------|--------------|-------|--------------|-------|------------|
| A+   |  100 to 96.5 | A     | 96.4 to 93.5 | A-    | 93.4 to 90 |
| B+   | 89.9 to 86.5 | B     | 86.4 to 83.5 | B-    | 83.4 to 80 |
| C+   | 79.9 to 76.5 | C     | 76.4 to 73.5 | C-    | 73.4 to 70 |
| D+   | 69.9 to 66.5 | D     | 66.4 to 63.5 | D-    | 63.4 to 60 |
| F    | 59.9 to 0    |       |              |       |            |

---

## Discussion sections

- Try to go only to the discussion section that you are registered in, 
otherwise there might not be enough space. 

--

- If at any point you can't make it to your discussion section and would like
to go to another consult it with the Ta assigned to your section and the section
that you plan to attend via email.

--

- All sections are in ALP 3600

---

## How to Contact us

- The best way to contact us is via email, we will try to respond within the 
following 24 hrs. 

--

- Try to look at the homework and see if you have any questions. We can't 
promise that we will be able to respond in time if the deadline is close.

--

- The syllabus is on Canvas, make sure to read it.

---

class: inverse, middle, center

# Review R

---

## RStudio and RStudio cloud

- There are two ways in which you can get R and RStudio for this class. 

--

1. Install [R](https://cran.microsoft.com/) and [Rstudio](https://www.rstudio.com/products/rstudio/download/#download) on your 
desktop/laptop.
  - First download and install R and then RStudio.

1. Start an account in [RStudio Cloud](https://rstudio.cloud/).

--

- Both methods are free!

---

## RStudio

- RStudio is a graphical interface for R. 

- The window is divided in four parts.

--

## Set up

- If you have installed R and RStudio in your personal computer then you will 
need a version of LaTex in order to generate the `.pdf` file that you have
to submit for your homework.

--

- Copy, paste and then run the following two lines in your RStudio console

```{r install-tiny, echo = TRUE, eval = FALSE}
# install tinytex package
install.packages('tinytex')

#install tinytex to be able to generate pdf files
tinytex::install_tinytex()
```

- You only need to do this once!

---

## Tidyverse

- For this class we will use many functions that are not part of base R but
that are in a separate package that can be downloaded. 

--

- To download and install the [tidyverse](https://www.tidyverse.org/packages/)
package you can copy and paste the following line of code into RStudio

```{r install-tidy, echo = TRUE, eval = FALSE}
# install tidyverse core packages
install.packages("tidyverse")
```

- Again you only need to do this once!

--

- packages in R are code that other people have written and made available. This
code takes the form of functions that we can use after downloading the package. 

--

- Many people use `tidyverse` so if you ever have a question and you can't 
reach us, you can Google it! 

- I promise someone else has asked that question and there will be a good 
answer.
