---
title: "Lecture 21"
subtitle: "Linear Regression"
author: "Psych 10 C"
institute: "University of California, Irvine"
date: "05/18/2022"
output:
  xaringan::moon_reader:
    self_contained: true
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-extra, echo = FALSE}
xaringanExtra::use_tile_view()

xaringanExtra::use_fit_screen()

xaringanExtra::use_editable(expires = 1)

xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         #<<
  mute_unhighlighted_code = TRUE  #<<
)

htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

```{r load-tidy, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(flipbookr)

knitr::opts_chunk$set(fig.width = 6, message = FALSE, 
                      warning = FALSE, comment = "", 
                      cache = F)

pressure <- read_csv(file = here::here("data/week-8/sys-pressure.csv"))
```

## Linear regression

- Last class we finished our example with multiple linear regression in a mental
rotation task.

--

- Today we will look at another example, however, this time we will not have a 
hypothesis, so we will have to take a "brute force" approach.

--

- We are interested in studying the effects of age and height on blood 
pressure.

--

- We are not sure if only one or both of this variables are good predictors so 
we want to compare all the models that we can build using these two variables.

--

- This will be similar to the notion of main effects and interactions in 
factorial designs.

---

## Data

- Now that we have a research question (are height and age good predictors of 
blood pressure), we need to look at the data in the study.

--

- There are 50 participants in this study, all of whom had their blood pressure 
taken during a routine check up. The average blood pressure of participants was
`r mean(pressure$blood_pressure)` (mmHg), with a range
`r min(pressure$blood_pressure)` to `r max(pressure$blood_pressure)`.

--

- The age of the participants ranged from `r min(pressure$age)` to 
`r max(pressure$age)`, with an average of `r mean(pressure$age)` years.

--

- The height of participants ranged from `r min(pressure$height)` to 
`r max(pressure$height)` with an average of `r mean(pressure$height)` inches.

--

- From those participants in the study 25 are female and the rest are male.

---

## Models

- Now that we have a description of the data we can visualize our observations
using a scatter plot, in this case we are interested in two variables (age and
height) so we can make two independent graphs.

--

.pull-left[
```{r blood-age, fig.align='center', echo=FALSE, fig.width=8}
color_sex <- c("#ff6361", "#58508d")
ggplot(data = pressure) + 
  aes(x = age) + 
  aes(y = blood_pressure) + 
  geom_point(size = 5, pch = 21, fill = color_sex[1], col = "white") + 
  theme_classic() + 
  xlab("Participant age") + 
  ylab("Blood pressure") + 
  theme(axis.title.x = element_text(size = 30),
        axis.title.y = element_text(size = 30))
```
]

.pull-right[
```{r blood-height, fig.align='center', echo=FALSE, fig.width=8}
ggplot(data = pressure) + 
  aes(x = height) + 
  aes(y = blood_pressure) +
  geom_point(size = 5, pch = 21, fill = color_sex[1], col = "white") + 
  theme_classic() + 
  xlab("Participant height") + 
  ylab("Blood pressure") + 
  theme(axis.title.x = element_text(size = 30),
        axis.title.y = element_text(size = 30))
```
]










