---
title: "Lecture 7"
subtitle: "Paired Samples"
author: "Psych 10 C"
institute: "University of California, Irvine"
date: "04/13/2022"
output:
  xaringan::moon_reader:
    self_contained: true
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-extra, echo = FALSE}
xaringanExtra::use_tile_view()

xaringanExtra::use_fit_screen()

xaringanExtra::use_editable(expires = 1)

xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         #<<
  mute_unhighlighted_code = TRUE  #<<
)

htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

```{r load-tidy, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)

library(flipbookr)

knitr::opts_chunk$set(fig.width = 6, message = FALSE, 
                      warning = FALSE, comment = "", 
                      cache = F)
link <- "https://raw.githubusercontent.com/ManuelVU/psych-10c-data/main/example-smoke.csv"
smokers <- read_csv(file = link)
```

## Paired samples

- Until now we have been working with an experimental design known as
**between subjects** design.

--

- This means that each participant in our sample can only be included in one 
group at a time.

--

- For example, in the smokers data, participant can be either on the smokers 
group or on the non-smokers group but they can't be in two.

--

- We call this a **between subjects** design because the data of one group 
 of participants is compared with data from a different group of participants

--

- However, we could have a design where each participant can respond in our 
two groups. 

---

# Example

- **Problem:** We are at the end of the semester at a university and we want to
know if our students have improved their grades. 

--

- There is variability in grades but we wold like to make a generalization, and
say that either they improved during the semester or that they didn't 
(as a group).

--

- To solve this problem we can use the students' scores on the midterm and their
scores on the final.

--

- However, notice that the observations between the groups will be correlated. 
This is a problem for our approach, as we stated in our model that the 
observations that we had where independent!

--

- The correlation in our observations comes from the fact that we are measuring
the same participant twice. The student 1 midterm and final score will have some
degree of correlation because they come from the same person.

--

- In order to avoid the problem of correlations between our two observations we
need to get reed of the "redundant" information.

---

## Example

- A simple way to get rid of the correlation is by just taking the difference 
between the two scores (this will only work when the correlation comes from 
us having two observations of the same participant).

--

- 


