---
title: "Lecture 11"
subtitle: "Factorial Designs"
author: "Psych 10 C"
institute: "University of California, Irvine"
date: "04/22/2022"
output:
  xaringan::moon_reader:
    self_contained: true
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-extra, echo = FALSE}
xaringanExtra::use_tile_view()

xaringanExtra::use_fit_screen()

xaringanExtra::use_editable(expires = 1)

xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         #<<
  mute_unhighlighted_code = TRUE  #<<
)

htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

```{r load-tidy, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)

knitr::opts_chunk$set(fig.width = 6, message = FALSE, 
                      warning = FALSE, comment = "", 
                      cache = F)
link <- "https://raw.githubusercontent.com/ManuelVU/psych-10c-data/main/anxiety-1-ex.csv"
anxiety <- read_csv(file = link) %>% 
  mutate("id" = as.character(id),
         "anxiety" = as.integer(anxiety),
         "cohort" = as.character(cohort))
```

## Factorial designs

- To this day we have talked abnout comparisons between groups difined by the 
values of single variable.

--

- We had comparisons between two groups when our independent variable could take
two values and participants where assigned at random to either of those values.

--

- We also had comparisons when we had two measures of the same participant, 
these designs are known as paired samples and our objective was to test the 
differences between the two measures.

--

- A third problem was when we had multiple values for an independent variable, 
for example, when we have students that belong to multiple cohorts or when 
we have multiple conditions in an experiment.

--

- One importat thing to note is that, even when our independen variable can take 
more than two values we are still talking about a single independent variable.

---

## Factorial designs

- In contrast to the methods that we have talked about in class, there are 
problems that will require us to look at the effects of more than one variable
at a time. 

--

- For example, when we test a new drug designed as a treatment to reduce blood 
pressure, it might be important to consider if the participant is male or female.

--

- The key point is that sometimes in our experiments we would like to consider 
effects of more than one variable at a time, and how those effecs migh interact 
to give rise to our observations.

--

- When we have more than one independent variable in an experiment which can 
take 2 or more categorical values (like smoking status, cohort, treatment, etc.)
we say call the experiment a **Factorial design**.

---

## Factorial designs 

- Whenever we have two or more independent variables that can take 2 or more 
categorical values we refer to them as **factors**. For example: 

--

 - In our study about the relation between smoking status and lung capacity we 
  can refer to smoking status as:  "the factor smoking".
  
--

  - In our study about the effects of a new drug on blood preassure we can refer
  to the treatment condition as: "the factor treatment".
  
--

- In order to describe our experimental designs we use a type of notation 
designed specifically for these kind of problems.

---

## Factorial designs

- For example, we are interested in comparing the performance of participants
in a recognition memory task versus a free recall task using high frequency
words and low frequency words.

--

- In this example, we have two factors, the firs one is the task factor, 
the second one is the word frequency factor.

--

- Furthermore, each factor can take two different values, the task factor
can be recognition or free recall and the word frequency factor can be high or 
low frequency.

--

- Then we would call this experimental design a $2\times 2$ **factorial design**, 
where the first number refers to the number of levels (values) of the first 
factor and the second refers to the levels (values) of the second factor.

---

## Factorial designs and participant assignment

- If participants are assigned to a single combination of the values of the 
factors, for example, one group of participants only responds to the free recall 
tests with high frequency words and a different group responds to the free recall
test with low frequency words, and so on. We refer to the design as a $2\times 2$ 
**between subjects factorial design**.

--

- If all participants respond to **all** combinations of the factors, we refer 
to it as a $2\times 2$ **within subjects factorial design**.

--

- This is very similar to what we had in the case of a single independent 
variable. 

--

  - A between subjects design is when different groups of participants are 
assigned to different values of the independent variable. 

--

  - A within subjects design was when participants responded to both levels of 
  our independent variable (like the before and after training problem in HW 2).

---

## Factorial designs and participant assignment.

- Now that we have more than one independent variable, we can have a new 
experimental design, **Mixed designs**.

--

- A **Mixed design** refers to an experiment where one of our factors 
(independent variables) is treated as between subjects and another is treated as
within subjects.

--

- For example, if we have the same group of participants responding to our free
recall test with both low and high frequency words and we have a **different** 
group of participants responding to the recognition memory task with both 
low and high frequency words.

- We will call this a $2\times2$ **mixed factorial design**.

---

## Mixed designs

- Whenever we have a mixed design, we have to specify which factors are treated
as between and which factors are treated as within subjects.

--

- From our previous example, an accurate description of the design would be:

--

  - This is a $2\times2$ **mixed factorial design** where the test factor was 
  implemented between subjects (different groups do different tasks) and the 
  word frequency factor was implemented within subjects (every participant 
  responded to the task with both high and low frequency words).
  
--

- An important thing to notice is that each type of design will have a different
number of independent groups, and that the number of levels that each group
would look at during the experiment will also be different.

---

## Factorial designs

- Using our memory experiment with 2 levels of each factor (independent 
variables).

--

  - If we had a $2\times2$ **between subjects factorial design**, that means 
  that we will have 4 groups and that each group will look at a single 
  combination of the levels of the factors (*e.g.* free recall - high frequency,
  free recall - low frequency, recognition - high frequency, 
  recognition - low frequency).
  
--

  - If we had a $2\times2$ **within subjects factorial design**, that means that
  we have a single group and that all participants in that group look at every
  combination of the levels of our two factors.
  
--

  - Finally, ff we had a $2\times2$ **mixed factorial design** with task 
  manipulated between subjects and word frequency manipulated within subjects. 
  We would have 2 groups (one performing a free recall tasks and the other a
  recognition tasks) and each group would see two values of our second
  factor (each group would do the task with low and high frequency words 
  independently).
  
---

## Factorial designs 

We have a $3\times 2$ between subjects design,

.can-edit.key-likes[
How many factors do we have in the experiment?

- **ANS:**

How many levels does each factor have?

- **ANS:**

How many independent groups do we have on the experiment?

- **ANS:**

How many combinations of the factors levels does each group see?

- **ANS:**

]

---

## Factorial designs 

We have a $2\times 2\times 3$ within subjects design,

.can-edit.key-likes[
How many factors do we have in the experiment?

- **ANS:**

How many levels does each factor have?

- **ANS:**

How many independent groups do we have on the experiment?

- **ANS:**

How many combinations of the factors levels does each group see?

- **ANS:**

]

---

## Factorial designs 

We have a $4\times 2$ mixed design, with the first factor manipulated between 
subjects and the second factor manipulated within subjects:

.can-edit.key-likes[
How many levels does each factor have?

- **ANS:**

How many independent groups do we have on the experiment?

- **ANS:**

How many combinations of the factors levels does each group see?

- **ANS:**

]

---

## Between subjects factorial designs

- Within subjects and mixed designs require more "sophisticated" methods to be 
analyzed than what we can cover in a 6 weeks. Therefore, we will focus on 
between subjects factorial designs.




