---
title: "Lecture 2"
subtitle: "New variales and Plots"
author: "Psych 10 C"
institute: "University of California, Irvine"
date: "05/28/2022"
output:
  xaringan::moon_reader:
    self_contained: true
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-extra, echo = FALSE}
xaringanExtra::use_tile_view()

xaringanExtra::use_fit_screen()

xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         #<<
  mute_unhighlighted_code = TRUE  #<<
)

htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

```{r load-tidy, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(flipbookr)
knitr::opts_chunk$set(fig.width = 6, message = FALSE, 
                      warning = FALSE, comment = "", 
                      cache = F)
link <- "https://raw.githubusercontent.com/ManuelVU/psych-10c-data/main/example.csv"
memory <- read_csv(file = link)
```

## Load data into R

- We will keep working with the memory data from last class

```{r data-example, echo = FALSE}
DT::datatable(
  head(x = memory, n = 8),
  fillContainer = FALSE, options = list(pageLength = 8)
)
```

---

## Creating new variables

- Sometimes we will like to work with some transformation of the variables
that we have on a data file. 

--

- For example we could want to have a variable that tells us if the number 
of correctly recalled words was from the first or second test.

--

This can make some plots easier to make!

---

## Creating a new variable

- We will create a new variable that takes the value "test-1" if the test was 
perforned after 300 seconds of study and takes the value "test-2" if it was 
3600 seconds after:

```{r id-test, echo = TRUE}
memory <- memory %>% 
  mutate("test_id" = ifelse(test = time_test == 300, 
                            yes = "test_1", no = "test_2"))
head(x = memory, n = 4)
```

---

## Creating a new variable

- Using the **`mutate`** function we can create new variables using other functions 
in R (**`ifelse`** is a function in R)

--

- We will use our new variable to create plots of the variables that we are 
interested in.

---

class: inverse, center, middle

# Plots

## Histograms

---

## Histograms

- One of the ways in which we can visualize data is using histograms.

--

- A histogram represents a count of the number of times that a variable has 
appeared in our data.

---

```{r hist-code, include = FALSE}
ggplot(data = memory) + #BREAK
  aes(x = correct) + #BREAK
  aes(fill = test_id, color = test_id) + #BREAK
  geom_histogram(position="identity", 
                 binwidth = 1,  
                 alpha = 0.4) + #BREAK
  theme_classic() + #BREAK
  xlab("Number of correct recalls") + #BREAK
  ylab("Frequency") + #BREAK
  guides(fill = guide_legend("Test order"), color = "none") + #BREAK
  theme(axis.title.x = element_text(size = 20),
        axis.title.y = element_text(size = 20))
```

`r chunk_reveal(chunk_name = "hist-code", title = "### Histogram of correct recalls")`

---

## Histograms

- One of the main problems with histograms is that their shape depends on our
choice of the withd of the bars!

--

- A change on the shape can change our interpretation of the results so we need 
to be careful when making our choice.

--

- In general we can use histograms when we have a numeric variable that we want 
to visualize.