---
title: "Lecture 8"
subtitle: "Multiple Groups"
author: "Psych 10 C"
institute: "University of California, Irvine"
date: "04/15/2022"
output:
  xaringan::moon_reader:
    self_contained: true
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-extra, echo = FALSE}
xaringanExtra::use_tile_view()

xaringanExtra::use_fit_screen()

xaringanExtra::use_editable(expires = 1)

xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         #<<
  mute_unhighlighted_code = TRUE  #<<
)

htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

```{r load-tidy, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)

knitr::opts_chunk$set(fig.width = 6, message = FALSE, 
                      warning = FALSE, comment = "", 
                      cache = F)
anxiety <- read_csv(file = "~/Documents/git_projects/psych-10c/data/week-3/anxiety-1-ex.csv") %>% 
  mutate("id" = as.character(id),
         "anxiety" = as.integer(anxiety),
         "cohort" = as.character(cohort))
```

## Comparisons between two groups

- Let's look at another example of comparisons between two populations.

--

- First we need a research problem or question. 

--

- We are interested in studying the levels of anxiety in first year students at 
a university in two different cohorts, the first one started in 2018 and the 
second one in 2019.

--

- The University makes all students take a survey during the first week which 
includes a scale design to measure anxiety on a scale that goes from 0 to 20.

--

- We have been granted access to the data of 30 students of each cohort to 
analize if there are any differences between the levels of anxiety.

--

- Is this a paired samples (within subjects) design or an independent samples
(between subjects) design?

---

## First year anxiety

- Before we get the results, we want to formalize our models.

--

- **Null model:** There are no difference in anxiety levels between students in 
the 2018 cohort and students in the 2019 cohort. In other words, the anxiety 
level of each student is an independent sample of the distribution: 
$$y_{ij} \sim \text{Normal}(\mu,\sigma_0^2)$$ 
for $i=1,\dots,30$ students and $j = 1, 2$ where 1 represents that the student
belongs to the 2018 cohort and 2 represents students of the 2019 cohort.

--

- **Effects model:** The anxiety levels of students in the 2018 cohort are 
different from the levels of the 2019 cohort. In other words, the anxiety level
of a student in group $j=1,2$ where 1 denotes the 2018 cohort and 2 denotes 2019
are an independent sample of the distributions:
$$y_{ij} \sim \text{Normal}(\mu_j, \sigma_e^2)$$
for $i = 1,\dots,30$ students. 

---

## Data.

- Before we do any analysis, we ca look at our data:

```{r look-data-1, echo = FALSE}
DT::datatable(data = anxiety,
  fillContainer = FALSE, options = list(pageLength = 7))
```

---

## Visualizing data

- Now we can look at the distribution of anxiety scores by cohort using a 
histogram:

.pull-left[
```{r hist-anxiety, eval = FALSE}
ggplot(data = anxiety) + 
  aes(x = anxiety) + 
  aes(fill = cohort, color = cohort) + 
  geom_histogram(position="identity", 
                 binwidth = 1,  
                 alpha = 0.3) + 
  theme_classic() + 
  xlab("Anxiety score") + 
  ylab("Frequency") + 
  guides(fill = guide_legend("Cohort"), color = "none") + 
  theme(axis.title.x = element_text(size = 20),
        axis.title.y = element_text(size = 20))  
```
]
.pull-right[
```{r hist-anxiety-out, ref.label = "hist-anxiety", fig.align = 'center', echo = FALSE, fig.width = 5.7, fig.height = 5.3}

```
]

---

## Infering parameter values from observations

- Histogram doesn't show any systematic differences, however, to reach a 
conclusion we need to compare our two models:

--

**Null model:**
.pull-left[
```{r null-model}
anxiety <- anxiety %>% 
  mutate("null_pred" = round(mean(anxiety),3),
         "null_error" = round((anxiety - null_pred)^2,3))
```
]
.pull-right[
Prediction = `r anxiety$null_pred[1]`

SSE = `r round(sum(anxiety$null_error),3)`

Mean SE = `r round(1/nrow(anxiety) * sum(anxiety$null_error),3)`
]

--

**Effects model:**
.pull-left[
```{r eff-model}
group_means <- anxiety %>% 
  group_by(cohort) %>% 
  summarise("pred" = round(mean(anxiety),3))
anxiety <- anxiety %>% 
  mutate("eff_pred" = ifelse(test = cohort == "2018", yes = group_means$pred[1],
                             no = group_means$pred[2]),
         "eff_error" = round((anxiety - eff_pred)^2, 3))
```
]
.pull-right[
Prediction: 
- 2018 = `r group_means[1,2]`, 
- 2019 = `r group_means[2,2]` 

SSE = `r round(sum(anxiety$eff_error),3)`

Mean SE = `r round(1/nrow(anxiety) * sum(anxiety$eff_error),3)`
]

---

## Model evaluation

```{r errors, echo = FALSE}
sse_0 <- sum(anxiety$null_error)
sse_e <- sum(anxiety$eff_error)
mse_0 <- 1/nrow(anxiety) * sse_0
mse_e<- 1/nrow(anxiety) * sse_e
```

- The proportion of error accounted for by the Effects model was:

  - $R^2 = 1.5 \times 10^{-4}$ 
  
--

- In other words, the model that assumes that anxiety levels are different 
between the two cohorts explains `r 100*round((sse_0 - sse_e)/sse_0, 5)`% of the 
variability on anxiety levels.

--

- In comparison to previous examples, this is a small percentage of error that 
the effects model is accounting for.

---

## BIC

- The BIC associated to the Null model was equal to

$BIC_0 =$ `r round(nrow(anxiety) * log(mse_0) + log(nrow(anxiety)),3)`

--

- The BIC associated to the Effects model was:

$BIC_e =$ `r round(nrow(anxiety) * log(mse_e) + 2 * log(nrow(anxiety)),3)`

--

- Given that the BIC value for the Null model is lower than the BIC of the 
Effects model, we can conclude that:

  - The evidence suggests that there are no differences on anxiety levels 
  between First year students of the 2018 and 2019 cohorts.
  
---

class: inverse, middle, center

# Multiple Groups

---

## Comparing multiple groups

- On the previous example we saw that the evidence we had, suggested that there 
were no difference on anxiety levels between the two cohorts. However, what 
would happen if we took into account the 2020 cohort?

--

- When our independent variable takes more than two categorical values 
(e.g. multiple cohorts, multiple tests, etc.) we have to make some changes to 
our models.

--

- The **Null model** will remain the same, and again it formalizes the 
assumption that there are no differences between the groups.

--

- However, the **Effects model** now has to take into account the fact that
there are now more than 2 groups.

---

class: inverse, center, middle

# Effects Model

---

## Effects model

- Our new effects model will now formalize the assumption that **at least one**
of the groups is different.

--

- The problem now will be that we don't exactly now which one of the groups is 
different than the others, but for now, this is the best that we can do with
the two models that we have.

--

**Effects model:** Let $y_{ij}$ be the anxiety level of the *i-th* student from 
the *j-th* cohort, with $i = 1, \dots, 30$ and $j = 1, 2, 3$; where 1 represents
the 2018, 2 represents 2019 and 3 represents the 2020 cohort. Then, each 
observation is assumed to be an independent sample from one of 3 distributions:

$$y_{ij}\sim\text{Normal}(\mu_j,\sigma_e^2)$$

---

## Multiple groups: Predictions

- Now we have 4 parameters in total for the model, 3 expectations $(\mu_j)$ and 
1 error $\sigma_e^2$.

--

- Something that doesn't change is what our best guess for $\mu_j$ is. The 
estimator for the parameter $\mu_j$ will be the average of each group, except 
that now we have 3 groups for which to calculate the average.

--

- In our example the prediction for each cohort *j* can be written as:
$$\hat{\mu}_j = \frac{1}{n_j}\sum_{i=1}^{30}y_{ij}$$

--

- Where $n_j$ represents the total number of students in each cohort (in our 
example this number is the same for all cohorts, 30 students).

--

- In other words, our prediction about the anxiety levels $\hat{\mu}_j$ of 
students that belong to the *j-th* cohort will be the average of the *j-th* 
cohort.

---

## Multiple groups: Mean Squared Error

- Our best "guess" or estimator for the error of the Effects model will be 
similar to the one we had for the two groups case. 

--

- Again the only difference is that this time we have more than two groups or 
predictions.

--

- In our example about anxiety levels in 3 cohorts of First year students we 
have that:
$$\hat{\sigma}_e^2=\frac{1}{n}\sum_{j=1}^{3}\sum_{i=1}^{30}\left(y_{ij}-\hat{\mu}_j\right)^2$$

--

- Where this time $n$ represents the total number of students in our data, 
given that we have 3 cohorts each with 30 students, the total would be equal to 
$90$.

---

## Multiple groups: SSE, $R^2$ and BIC



