---
title: "Lecture 9"
subtitle: "Multiple Groups: Additional models"
author: "Psych 10 C"
institute: "University of California, Irvine"
date: "04/15/2022"
output:
  xaringan::moon_reader:
    self_contained: true
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-extra, echo = FALSE}
xaringanExtra::use_tile_view()

xaringanExtra::use_fit_screen()

xaringanExtra::use_editable(expires = 1)

xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         #<<
  mute_unhighlighted_code = TRUE  #<<
)

htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

```{r load-tidy, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)

knitr::opts_chunk$set(fig.width = 6, message = FALSE, 
                      warning = FALSE, comment = "", 
                      cache = F)  
link <- "https://raw.githubusercontent.com/ManuelVU/psych-10c-data/main/anxiety-2-ex.csv"
anxiety <- read_csv(file = link)
```

## Final Project

- As you already know, we will have a final project for this class instead of a 
final exam.

--

- This final project is worth 50% of your total grade and it should be done in 
groups of between 3 and 4 students.

--

- The objective is to have you write a short version of what a scientific paper 
should look like. Including the data analysis section.

--

- To make things "easier" for you, we have selected 3 problems with their 
corresponding data sets that you can choose from for your final project.

--

- To answer the research question of each project you will need a different 
approach, not all problems can be solved the same way. Therefore, before we 
look at the data we want you to choose one of the three problems.

---

### Problem 1:

- We want to know if people that have had a brain curgical procedure known as 
split brain, have problems identifing obhects presented at either side of their 
visual field.

--

### Problem 2:

- We want to know if the cognitive decline of a person from one year to the next 
is associated with their level of social interaction, physical activity and the 
fact that they used the app Luminosity during that year.

--

### Problem 3:

- We want to know how the asking price for an IKEA chair changes as a function
of the total cost of the matrials, if the person build the chaitr themselves and
the difficulty of building it.

---

## Final project

The paper must have the following points:

- Introduction: 4 to 5 sentences that summarize the problem, and why it's 
important. The idea is to convince your reader that what you are doing matters
you can use references to previous work (1 or 2 max).

--
  
- Methods: In one paragraph, explain what is in the data set and where it came 
from (how it was collected).

--
  
- Data: provide a summary of your data, you can use summary statistics like
the median, mean or variance. You can also use this section to highlight some
properties of the data that you are working on by using graphs.
  
---

## Final project

- Model comparison: Describe the models that you will be comparing using your 
current data and their main assumptions (does the model assume that groups are
equal?).
  
  - You should provide a summary of your analysis, a good way to do so can be
  to have a table that has the SSE of each model that you tested and the BIC
  associated to this model.

--

- Discussion: 

  - One paragraph on: What conclusions can you draw from the model comparisons
  about the experimental question.
  
--
  
  - One paragraph on: What might be the broader implications of this finding?
  
--
  
  - One paragraph on: What are the limitations of this study and/or what are the 
  gaps left by this study? Be sure to explain why the limitation could impact 
  the results.
  
---

class: inverse, center, middle

# Review of BIC's

---

## BIC

- Last week we mentioned that regardless of what data we have, a model that 
assumes that there is more that one distribution will always be better in 
terms of the error it produces in comparison to a simpler model.

--

- We can try to understand this intuitively by looking a graph that uses 
a single mean as a prediction. To do this we will use our smokers data:

--

```{r, echo = FALSE, fig.align='center', fig.height = 6}
link <- "https://raw.githubusercontent.com/ManuelVU/psych-10c-data/main/example-smoke.csv"
smokers <- read_csv(file = link)
par(mai = c(0.6,0.6,0.6,0.6))
par(oma = c(3, 3, 0.2, 0.2))
plot(x = seq(1,nrow(smokers)), y = sort(smokers$lung_capacity, decreasing = FALSE),
     axes = FALSE, ann = FALSE, col = rep(c("#c80064","#54bebe"), each = 4),
     pch = rep(c(15,19), each = 4), cex = 1.6)
box(bty = "l", lwd = 1.3)
axis(2, las = 2)
mtext(text = "Lung capacity", line = 2.3, cex = 1.6, side = 2)
mtext(text = "Smokers", line = 1, cex = 1.8, side = 1, at = 2.5)
mtext(text = "Non-smokers", line = 1, cex = 1.8, side = 1, at = 6.5)
```
