---
title: "Lecture 3"
subtitle: "Summary Statistics"
author: "Psych 10 C"
institute: "University of California, Irvine"
date: "03/30/2022"
output:
  xaringan::moon_reader:
    self_contained: true
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-extra, echo = FALSE}
xaringanExtra::use_tile_view()

xaringanExtra::use_fit_screen()

xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         #<<
  mute_unhighlighted_code = TRUE  #<<
)

htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

```{r load-tidy, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(flipbookr)
knitr::opts_chunk$set(fig.width = 6, message = FALSE, 
                      warning = FALSE, comment = "", 
                      cache = F)
link <- "https://raw.githubusercontent.com/ManuelVU/psych-10c-data/main/example.csv"
memory <- read_csv(file = link)
```

## Summary Statistics

- Another common way in which information from an experiment or a survey can be 
summarized is by using a statistic.

--

- Statistics are **functions** of **random variables** in an experiment 
that can be used to convey information about the location of our observations 
and how they vary.

--

- Not all the variables in an experiment are equally important, so we don't 
usually look for ways to visualize them, but we still want to make sure that 
we give all the information we can about our sample.

--

- In that case we can use summary statistics to report some of their properties.

---

class: inverse, center, middle

# Random Variables and Functions

---

## Random variables

- Statisticians are very bad at naming things ...

--

- So when you think of the words random and variable what comes to mind first?

--

- The formal definition is the opposite!

--

- **Definition:** A random variable is a function of the outcomes of an 
experiment.

---

# Functions

- Functions have formal definitions, however, what we have to remember is 
how they "work".

--

- Intuitively, we can think of functions as rules about how to associate two 
groups of "things".

--

- For example, if throw a coin and then record if it landed as heads or tails. 
A function could be a rule that states:

  - $x = 0$ If the outcome is tails and $x = 1$ if the outcome is heads. 

--

- This is a simple rule that just let's us assign numbers to a variable $x$ 
depending on the result of a coin toss. In other words, $x$ is a function of the 
outcome of the experiment.

--

- Another simple function would be $y = x + 1$. 

  - This function, tells us that whatever the value of $x$ is, we can 
  get the value of $y$ by adding $1$ to $x$.

--

- Regardless of how complex they look, we can think of functions as a "map"
that specifies how we get from the values of one variable to the values of 
another.

---

## Back to Random Variables

- Random variables are neither random nor variables, they are simply the rules 
we use to assign numbers to the outcomes of an experiment. 

--

- In other words they are deterministic functions (see statisticians are really 
bad at naming).

--

- In our previous example with the coin toss, $x$ can be considered a random 
variable, as it is a rule that allows us to assign a numeric value (0 or 1) to 
the outcome of the experiment (heads or tails).

---

## Example with the memory experiment.

- Let's think of our memory experiment. For a given participant and test (say 
participant 1, test 1), each time we present a word it can be the case that the 
participant response was that the word was on the original list or not.

--

- Additionally, each word that we present was either on the test or it wasn't.

--

- We can treat our participants responses as probabilistic given that we are not 
sure how any person would respond to any word. 

--

- Now we can create a random variable that says:

--

  - if the word was on the original list **and** the participant responds that 
  the word was on the original list, then $x = 1$.

--

  - if the word was in the original list **and** the participant responds that 
  the word was **not** in the original list, then $x = 0$

--

- If we record the value of $x$ for all trials for a given participant we would
have 50 different values that indicate, for each word presented whether that 
particular response was correct or not.

---

## Statistics

- The examples we have talked about are all Statistics, a statistic is just
a function of our sample (data).

--

- In our memory experiment we don't have the record of the responses of each 
participant to each word, we have something "simpler".

--

- We have another random variable that adds all the correct responses.

--

- We have lost some information while doing so. Can you guess what information 
has been lost?

--

- We made a trade between the information about the order of the correct 
responses in exchange for a summary of the experiment, the total number of 
correct responses.

---

## Statistics

- Every time we use a statistic (function of our experimental outcomes) we 
either:

--

  - keep the same information (for example assigning a value of 1 to a heads 
  in the coin toss) 

--

  - Lose information (for example when we take the number of correct responses
  in the memory experiment) 

--

- In the majority of the examples of this course this will not be a problem, 
however, it is important to keep this loss of information in mind.

---

class: inverse, center, middle

# Commonly Used Statistics

## The Mean

---

## Mean

- As you know from Psych 10 B, one of the properties of a r.v. that we are 
interested in is its expected value.

--

- This value can be calculated with the formula: 
$$\mathbb{E}(x) = \sum_x x \ p(x)$$

--

- We are faced with a problem here as, when we gather data form an experiment, 
we don't know the probability of each of the values of our random variable.

--

- In other words we don't know $p(x)$.

--

- For example, what is the probability that a participant has 40 correct 
responses?

---

## Mean

- Fortunately, we can mathematically prove that the **average** of a random 
variable will be close to the expected value.

--

- This is true regardless of the values of $p(x)$.

--

- Of course this is just an approximation and will therefore be prone to error.

--

- But it will be our best guess!



