---
title: "Review R"
subtitle: ""
author: "Psych 10 C"
institute: "University of California, Irvine"
date: "04/04/2022"
output:
  xaringan::moon_reader:
    self_contained: true
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-extra, echo = FALSE}
xaringanExtra::use_tile_view()

xaringanExtra::use_fit_screen()

xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         #<<
  mute_unhighlighted_code = TRUE  #<<
)

htmltools::tagList(
  xaringanExtra::use_clipboard(
    button_text = "<i class=\"fa fa-clipboard\"></i>",
    success_text = "<i class=\"fa fa-check\" style=\"color: #90BE6D\"></i>",
    error_text = "<i class=\"fa fa-times-circle\" style=\"color: #F94144\"></i>"
  ),
  rmarkdown::html_dependency_font_awesome()
)
```

```{r load-tidy, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)

library(flipbookr)

knitr::opts_chunk$set(fig.width = 6, message = FALSE, 
                      warning = FALSE, comment = "", 
                      cache = F)
```

## Review of R 

- In this class we will review some of the basic elements in R.

--

- R is a programming language

--

- One of its main advantages is that users are able to write code and functions
and make them available for other users. 

--

- These functions come in the form of packages which we can download using
RStudio.

---

class: inverse, center, middle

# Scripts and R Markdown

---

## R scripts

- R scripts are used to write and store code, for example, if we want to analyze
the data of an experiment, we can write a script that does multiple operations
and stores new data set with what we want to show.

--

- We use scripts to make sure that we can replicate a data analysis, or to 
save the code we used to generate a plot.

---

## R Markdown

- R Markdown files `.rmd` are different from simple R scripts.

--

- They allow us to integrate text and R code into a single file.

--

- Whenever we "knit" this type of files, the integrated R code is executed and 
the variables are saved momentarily.

--

- Notice that the variables are **not** saved in R as it would if we were
executing or "running" an R script. On the contrary, the variables are just 
saved in memory while the file is being created, which allows us to use 
them inside the document.

--

- This is really useful when we have graphs or outcomes of a data analysis that 
we want to integrate into a report. 

---

class: inverse, center, middle

# RStudio

---

## RStudio

- RStudio is a graphical interface for R. It makes it easier to interact with R
by maintaining the console, scripts and graphs all in the same place.

--

- From this point on, we will work simultaneously with RStudio and the lecture 
notes as you should do when working on your assignments.

--

- Let's start by working with an R markdown file first

--

Let's say that we have been assigned a data file to work with on a homework, 
and that we are given a URL to the file such as:

```{r link-example, eval = FALSE}
link <- "https://raw.githubusercontent.com/ManuelVU/psych-10c-data/main/stroop.csv"
```



