---
title: "Homework 2"
author: "Write your name here"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r load_libraries, echo = FALSE}
# Code 0: Loading the libraries
#
# Use this chunk to load whatever libraries you need. They should already have been
# installed for the first homework, but you need to load them using library()
# in order to use them. IMPORTANT: Remember to delete the comments inside the code
# chunks BEFORE submitting your homework.
```

## Section 1

**Background:** For this section, we will work with our coffee experiment. 
This data set comes from an experiment designed to test whether drinking coffee 
makes people smarter. Participants were divided into two groups. All 
participants took an IQ test, but participants in **group 1** 
drank water 5 minutes before the test, while participants in **group 2** drank 
coffee. (**The data is not the same as in Homework 1**)

We have two competing hypotheses (models). Model 1 states that drinking 
coffee doesn't make a difference in IQ test scores between people who 
drank water (group 1) and people who drank coffee (group 2). Model 2 predicts 
that people who drank coffee will be smarter; in other words, the IQ 
test scores should be different between groups (with group 2 being better than 
group 1). Both hypotheses assume that variances are equal across observations 
and groups.

```{r load_data, echo = TRUE}
# Code 1: Load the data (Hint: Go back to Lec-4)
#
# In this chunk you should write (a.k.a. copy from the slides) the code needed 
# to load the data for section 1.
```

1. Make a density graph of Model 1 (**no differences between groups**). Remember 
that at this point, the actual values of x don't matter. We are
interested in **what the model expects the data to look like**. Do it with a 
variance equal to 3.

```{r density-m1-s1, echo = TRUE, fig.align = 'center'}
# Code 2: Density plot for Model 1
#
# In this code chunk you can write the code used to generate the density plot
# for Model 1. Remember that the values on the x axis are not important. We 
# just want to visualize what the model expects the data to look like.
```

2. Obtain the predictions of Model 1 add the required values as a 
new variable (column) to the data set (remember to give this new variable a 
name!). What are the values of IQ predicted by Model 1 for each group?

```{r pred-m1-s1, echo = TRUE}
# Code 3: Testing the predictions made by Model 1
#
# In this code chunk you can write the functions that you need to
# calculate the predictions of model 1 and add it to the data set as a new
# column. Report the value(s) as text in the space below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

3. Calculate the squared error (SE) of each observation under Model 1 by adding
a new column to our data set. What is the mean Sum of Squared Errors 
(m-SSE) of the "no group differences" model (model 1)?

```{r sse-m1-s1, echo = TRUE}
# Code 4: Add a Squared Error column and compute the m-SSE for Model 1
# 
# In this code chunk you can write the functions that you need to use in order 
# to calculate the squared error of each observation and the functions you need 
# to calculate the mean sum of squared errors of the "no differences" model.
```

4. What is the value of the Bayesian Information Criterion (BIC) for Model 1?

```{r bic-m1-s1, echo = TRUE}
# Code 5: Calculate the BIC for Model 1
#
# In this code chunk you can write the functions that you need in order to 
# calculate the BIC associated with Model 1. Save it as a new variable and 
# report the value as text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

5. Make a density graph of Model 2 (**coffee group is better**). Remember that 
the actual values of x don't matter. Right now, we only want to visualize
what the model expects the data to look like. Assume that the model variance 
is equal to 3.

```{r density-m2-s1, echo = TRUE, fig.align = 'center'}
# Code 6: Density plot for Model 2
#
# In this code chunk you can write the code used to generate the density plot
# for Model 2. Remember that the values on the x axis are not important. What we 
# are looking at is what the model expects the data to look like.
```

6. Obtain the predictions of Model 2 add the required values as a new variable 
(column) to the data set (remember to give this new variable a DIFFERENT name!). 
What are the values of IQ predicted by Model 2 for each group?

```{r pred-m2-s1, echo = TRUE}
# Code 7: Testing the predictions made by Model 2
#
# In this code chunk you can write the functions that you need to use in order 
# to calculate the predictions of Model 2 and add it to the data set as a new
# column. Report the value(s) as text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

7. Calculate the squared error (SE) for each observation under Model 2 by adding
a new column to the data set (again, watch out the names you give to the new 
column!). What is the mean Sum of Squared Errors (m-SSE) of this "group 
differences" model (Model 2)?

```{r sse-m2-s1, echo = TRUE}
# Code 8: Add a Squared Error column and compute the m-SSE for Model 1
#
# In this code chunk you can write the functions that you need to use in order 
# to calculate the squared error of each observation and the functions you need 
# to calculate the mean sum of squared errors of the "group differences" model.
```

**ANS:** [WRITE YOUR ANSWER HERE]

8. What is the value of the Bayesian Information Criterion (BIC) for Model 2?

```{r bic-m2-s1, echo = TRUE}
# Code 9: Calculate the BIC for Model 1
#
# In this code chunk you can write the functions that you need in order to 
# calculate the BIC associated with Model 2. Save it as a new variable and 
# report the value as text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

9. Which of the two models would you select and why?

**ANS:** [WRITE YOUR ANSWER HERE]

10. What does the Model that you selected tell us about our research question?

**ANS:** [WRITE YOUR ANSWER HERE]


## Section 2

**Background:** For this section, we will work with our response time experiment. 
This data set comes from an experiment designed to test whether a new 
training technique can improve participants' ability to navigate a maze. 
Participants first completed a maze after playing a game that lasted as long as 
the new training approach. Then they underwent the new training and completed a 
second maze of the same difficulty (we will assume that, if there is a difference 
in response times, it is only because of the training and not because of experience 
with maze solving). The data provides information about the time (in seconds) 
in which each participant completed each maze. 
(**The data is not the same as in Homework 1**)

Once again, we have two competing models (hypotheses) about the performance of 
participants on the maze before and after training. Model 1 (**no improvement**)
states that the new training method wouldn't make any difference. In other 
words, Model 1 predicts that there will be no improvement in terms of how 
quickly people solve the maze after being trained. 

Model 2 (**improvement**) states that the new training method should be
effective, and so the time people need to solve the maze should be shorter 
the second time (after participants have undergone the training).

```{r load_data2, echo = TRUE}
# Code 10: Load the data (Hint: Go back to Lec-4 and use the link)
#
# In this section you should write the code needed to load the data for
# section 2. IMPORTANT: Remember to delete the comments inside the code 
# chunks BEFORE submitting your homework.
```

11. Make a density graph for Model 1 (**no improvement**). To make the plot, 
assume that the model variance is equal to 2.

```{r density-m1-s2, echo = TRUE, fig.align = 'center'}
# Code 11: Density plot for Model 1
#
# In this code chunk you can write the code used to generate the density plot
# for model 1.
```

12. What are the predictions made by Model 1 (**no improvement**)? Again, make 
sure to add the corresponding values as a new variable (column) to our data base.

```{r pred-m1-s2, echo = TRUE}
# Code 12: Testing the predictions made by Model 1
#
# In this code chunk you can write the functions that you need to use in order 
# to calculate the predictions of model 1 and add it to the data set as a new
# column. Report the value(s) as text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

13. Calculate the squared error (SE) of each observation for Model 1 and
add them to the data set as a new variable. What is the mean Sum of Squared 
Errors (m-SSE) of the "no improvement" model (Model 1)?

```{r sse-m1-s2, echo = TRUE}
# Code 13: Adding a column for the SE and calculating the m-SSE
#
# In this code chunk you can write the functions that you need to use in order 
# to calculate the squared error of each observation and the functions you need 
# to calculate the sum of squared errors of the "no improvement" model.
```
**ANS:** [WRITE YOUR ANSWER HERE]

14. What is the value of the Bayesian Information Criterion for Model 1?

```{r bic-m1-s2, echo = TRUE}
# Code 14: Calculating the BIC for Model 1
#
# In this code chunk you can write the functions that you need in order to 
# calculate the BIC associated to model 1. Save it as a new variable and 
# report the value as text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

15. Make a density graph for Model 2 (**improvement**). Remember that at 
this point, the units (values of x) don't matter. We are only interested in what 
the model expects the data to look like.

```{r density-m2-s2, echo = TRUE, fig.align = 'center'}
# Code 15: Density plot for Model 2
#
# In this code chunk you can write the code used to generate the density plot
# for model 2. Remember that the values on the x axis are not important, what we 
# look at is what the model expects the data to look like.
```

16. Obtain the predictions of the second model and add them as a new variable 
(column) to the data set (remember to give the variable a DIFFERENT name!). What 
are the predictions of Model 2?

```{r pred-m2-s2, echo = TRUE}
# Code 16: Testing the predictions of Model 2
#
# In this code chunk you can write the functions that you need to use in order 
# to calculate the predictions of model 2 and add it to the data set as a new
# column. Report the value(s) as text below
```

**ANS:** [WRITE YOUR ANSWER HERE]

17. Calculate the squared error (SE) of each observation for the second model 
and add them to the data set as a new variable. What is the mean Sum of Squared 
Errors (m-SSE) of the "improvement" model (model 2)?

```{r sse-m2-s2, echo = TRUE}
# Code 17: Adding a column for the SE and calculating the m-SSE
#
# In this code chunk you can write the functions that you need to use in order 
# to calculate the squared error of each observation and the functions you need 
# to calculate the mean sum of squared errors of the "improvement" model.
```

**ANS:** [WRITE YOUR ANSWER HERE]

18. What is the value of the Bayesian Information Criterion for model 2?

```{r bic-m2-s2, echo = TRUE}
# Code 18: Calculating the BIC for Model 2
#
# In this code chunk you can write the functions that you need in order to 
# calculate the BIC associated with model 2. Save it as a new variable and 
# report the value as text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

19. Which of the two models would you select and why?

**ANS:** [WRITE YOUR ANSWER HERE]


20. What does the Model that you selected tell us about our research question?

**ANS:** [WRITE YOUR ANSWER HERE]

