---
title: "Homework 2"
author: "Write your name here"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

## Section 1

**Background:** For section we will work with our coffee data set again. 
This data set comes from an experiment designed to test whether  drinking coffee 
makes people smarter. Participants were divided into  two groups. All 
participants took an IQ test, but participants in **group 1** 
drank water 5 minutes before the test, while participants in **group 2** drank 
coffee. 

We have two competing hypothesis (models), the first one is that 
drinking coffee doesn't make a difference in IQ test scores between people who 
drank water (group 1) and people who drank coffee (group 2). The second one 
states that people who drank coffee are smarter, in other words, that the IQ 
test scores should be different between groups. For both hypothesis we will 
assume that variances are equal across observations and groups.

```{r load_data, echo = TRUE}
# In this chunk you should write (a.k.a. copy from the slides) the code needed 
# to load the data for section 1. IMPORTANT: Remember to delete the comments 
# inside the code chunks BEFORE submitting your homework.
```

1. Write the first model (**no differences between groups**) in mathematical
notation

**ANS:** [WRITE YOUR ANSWER HERE]

2. Make a density graph of model 1 (**no differences between groups**). Remember 
that at this point the units (values of x) don't matter, we are only interested 
in what the model expects the data to look like.

```{r density-m1-s1, echo = TRUE, fig.align = 'center'}
# In this code chunk you can write the code used to generate the density plot
# for model 1. remember that the values on the x axis are not important, what we 
# look at is what the model expects the data to look like.
```

3. Calculate the prediction of the first model and add it as a new variable 
(column) to the data set (remember to give the variable a name!). What is the 
prediction of the first model?

```{r pred-m1-s1, echo = TRUE}
# In this code chunk you can write the functions that you need to use in order 
# to calculate the predictions of model one and add it to the data set as a new
# column. Report the value(s) as text text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

4. Calculate the squared error (SE) of each observation for the first model and
add it to the data set as a new variable. What is the mean Sum of Squared Errors 
(m-SSE) of the no group differences model (model 1)?

```{r sse-m1-s1, echo = TRUE}
# In this code chunk you can write the functions that you need to use in order 
# to calculate the squared error of each observation and the functions you need 
# to calculate the sum of squared errors of the no differences model.
```

5. What is the value of the Bayesian Information Criterion for model 1?

```{r bic-m1-s1, echo = TRUE}
# In this code chunk you can write the functions that you need in order to 
# calculate the BIC associated to model 1. Save it as a new variable and 
# report the value as text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

6. Write the second model (**groups are different**) in mathematical notation.

**ANS:** [WRITE YOUR ANSWER HERE]

7. Make a density graph of model 2 (**groups are different**). Remember that at 
this point the units (values of x) don't matter, we are only interested in what 
the model expects the data to look like.

```{r density-m2-s1, echo = TRUE, fig.align = 'center'}
# In this code chunk you can write the code used to generate the density plot
# for model 2. remember that the values on the x axis are not important, what we 
# look at is what the model expects the data to look like.
```

8. Calculate the prediction of the second model and add it as a new variable 
(column) to the data set (remember to give the variable a different name!). What 
are the prediction of the second model?

```{r pred-m2-s1, echo = TRUE}
# In this code chunk you can write the functions that you need to use in order 
# to calculate the predictions of model two and add it to the data set as a new
# column. Report the value(s) as text below
```

**ANS:** [WRITE YOUR ANSWER HERE]

9. Calculate the squared error (SE) of each observation for the second model and
add it to the data set as a new variable. What is the mean Sum of Squared Errors 
(m-SSE) of the group differences model (model 2)?

```{r sse-m2-s1, echo = TRUE}
# In this code chunk you can write the functions that you need to use in order 
# to calculate the squared error of each observation and the functions you need 
# to calculate the mean sum of squared errors of the group differences model.
```

**ANS:** [WRITE YOUR ANSWER HERE]

10. What is the value of the Bayesian Information Criterion for model 2?

```{r bic-m2-s1, echo = TRUE}
# In this code chunk you can write the functions that you need in order to 
# calculate the BIC associated to model 2. Save it as a new variable and 
# report the value as text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

11. Which of the two models would you select and why?

**ANS:** [WRITE YOUR ANSWER HERE]



## Section 2

**Background:** For section we will work with our response time data set again.
This data set comes from an experiment designed to test 
whether a new training technique can improve participants' ability to navigate 
a maze. Participants in the experimental condition underwent the new training, 
while participants in the control group played an unrelated game that lasted 
approximately as long as the training. Participants in both conditions were then 
instructed to complete a maze as quickly as possible. The data provides 
information about the time (in seconds) in which each participant completed the 
maze.

```{r load_data2, echo = TRUE}
# In this section you should write the code needed to load the data for
# section 2. IMPORTANT: Remember to delete the comments inside the code 
# chunks BEFORE submitting your homework.
```



