---
title: "Homework 1"
author: "Write your name here"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---
## Section 1

**Background:** This first data set comes from an experiment designed to test whether  drinking coffee makes people smarter. Participants were divided into two groups. All participants took an IQ test, but participants in **group 1** drank water 5 minutes before the test, while participants in **group 2** drank coffee.

```{r load_data, echo = TRUE}
# In this chunk you should write (a.k.a. copy from the slides) the code needed 
# to load the data for section 1. IMPORTANT: Remember to delete the comments 
# inside the code chunks BEFORE submitting your homework.
```

\small Questions 1 - 4 require you to use R code, questions 5 and 6 require you to draw conclusions from the output.

1. What is the average score on the *IQ* test for participants who **did not drink coffee?**

```{r mean_group1, echo = TRUE}
# In this code chunk you can write the function (or functions) that you need to 
# calculate the mean of the FIRST group, remember to save the result in a new 
# variable so that you can easily print it in text. Remember to delete 
# the comments inside the code chunks BEFORE submitting your homework.
```

**ANS:** [WRITE YOUR ANSWER HERE]

2. What is the average score on the *IQ* test for participants who **drink coffee?** 

```{r mean_group2, echo = TRUE}
# In this code chunk you can write the function (or functions) that you need to 
# calculate the mean of the SECOND group, remember to save the  result in a new 
# variable so that you can easily insert it in text. Remember to delete 
# the comments inside the code chunks BEFORE submitting your homework.
```

**ANS:** [WRITE YOUR ANSWER HERE]

3. Which group had more variation in their test scores on the *IQ* test, those who drink coffee or those who didn't? How do you know?

```{r variantion_group, echo = TRUE}
# In this code chink you can write the functions that you need to use in order 
# to calculate the variation of each group. Remember to assign each value to a 
# new variable so that you can use it in text.
```

**ANS:** [WRITE YOUR ANSWER HERE]

4. Create two histograms for the *IQ* test results that each group got (1 per group). In other words, one of these histograms must show the IQ scores for participants who were not given coffee before the IQ test, and the other histogram should show how participants who were given coffee performed. Describe what you see in the histograms in as few words as possible.

```{r histograms, echo = TRUE, fig.align = 'center'}
# Use this code chunk to make the histograms of the IQ test scores of each group
# it would be easier to compare results if you add transparency to the color! 
```

**ANS:** [WRITE YOUR ANSWER HERE]

5. We take the mean IQ of participants who drank coffee so that we can get a single number that helps describe the performance of all participants who drank coffee. Similarly, we take the mean IQ of participants who didn’t drink coffee so that we can get a single number that helps describe the behavior of participants who didn’t drink coffee. Which of these means do you think does a better job at describing the behavior of their respective group? Why?

**ANS:** [WRITE YOUR ANSWER HERE]

6. If someone asked you, does drinking coffee make a difference for how people perform on IQ tests, what would you say and why?

**ANS:** [WRITE YOUR ANSWER HERE]



## Section 2

**Background:** This second data set comes from an experiment designed to test whether a new training technique
can improve participants' ability to navigate a maze. Participants in the experimental condition underwent the 
new training, while participants in the control group played an unrelated game that lasted approximately as long as the 
training. Participants in both conditions were then instructed to complete a maze as quickly as possible. 
The data provides information about the time (in seconds) in which each participant completed the maze.

```{r load_data2, echo = TRUE}
# In this section you should write the code needed to load the data for
# section 2. IMPORTANT: Remember to delete the comments inside the code 
# chunks BEFORE submitting your homework.
```

\small Questions 7 - 10 require you to use R code, questions 11 - 13 require you to draw conclusions based on the outcome.

7. On average, how much did participants in the control condition improve their times? Make sure your answer is clear, both in terms of units and $+/-$.

```{r diff_group, echo = TRUE}
# In this section you should write the code needed to calculate the average 
# improvement for both groups, remember to save the result in a new variable and 
# to use it in the text accessing the elements of the vector.
```

**ANS:** [WRITE YOUR ANSWER HERE]

8. On average, how much did participants in the experimental condition improve their times? Again, make sure your answer is clear, both in terms of units and $+/-$.

**ANS:** [WRITE YOUR ANSWER HERE]

9. Which group had more variation in their improvement from one puzzle to the next, those who received the training or those who didn’t? How do you know?

```{r variantion_sec2, echo = TRUE}
# In this code chunk you can write the functions that you need to use in order 
# to calculate the variation of each group. Remember to assign each value to a 
# new variable so that you can use it in text.
```

**ANS:** [WRITE YOUR ANSWER HERE]

10. Create 2 boxplots. One should show how participants who weren’t given the 
training improved from the first puzzle to the second puzzle, and the other 
should show how participants who did get the training improved. These boxplots 
should be side by side for comparison purposes. Write a small comment on what 
you see.

```{r box-plots, echo = TRUE, fig.align = 'center'}
# Use this code chunk to make the boxplots of the average time to solve a maze 
# after training divided by group.
```

**ANS:** [WRITE YOUR ANSWER HERE]

11. Similar question 5 from Section 1, which mean do you think does a better job 
of describing participants behavior in each condition, the mean improvement of 
participants who did receive the training or the mean improvement of those who 
didn’t? Why?

**ANS:** [WRITE YOUR ANSWER HERE]

12. If someone asked you whether the training was effective in improving people’s performance on the logic puzzles, what would you say? Why?

**ANS:** [WRITE YOUR ANSWER HERE]

13. Considering the data you have looked at in this assignment, would you be more confident in saying that drinking coffee improves people’s performance on IQ tests or that the new training improves people’s performance on logic puzzles? Why?

**ANS:** [WRITE YOUR ANSWER HERE]