---
title: "homework 4"
author: "Manuel Villarreal"
date: "4/18/2022"
output: pdf_document
---

**Background:** For this section, we will work with the responses to a memory 
experiment. Researchers are interested in the effects of biological sex at birth 
and word frequency (how commonly a word is used in English) on the total number 
of words that can be correctly recalled from a list. For this experiment we have 
two factors (biological sex and word frequency) with two levels each (female vs. 
male and common vs. rare). For each combination of the factor levels we have 20 
observations for a total of 80 participants. For this problem we will consider a 
total of 4 models: (1) the Null, (2) biological sex main effect, (3) word 
frequency main effect and (4) the additive model. 

```{r load_data2, echo = TRUE}
# Code 14: Load the data (Hint: Go back to Lec-10 and use the link)
#
# In this section you should write the code needed to load the data for
# section 2. IMPORTANT: Remember to delete the comments inside the code 
# chunks BEFORE submitting your homework.
```

16. Obtain the grand mean and the effects of biological sex ($\hat{\alpha}$) and 
word frequency ($\hat{\beta}$).

```{r grand-eff-s2, echo = TRUE}
# Code 15: calculating grand mean and main effects.
#
# In this code chunk you can write the functions that you need to use in order 
# to calculate the grand mean and the effects of each factor.
```

**ANS:** [WRITE YOUR ANSWER HERE]

17. Obtain the predictions of the Null model and add them to the data. Next, 
calculate the error of each observation and add it as a new variable. Use those 
values to calculate the SSE and the Mean SE associated with the Null model. 
Report the prediction, SSE and Mean SE of the Null model below.

```{r null-m1-s2, echo = TRUE}
# Code 16: adding predictions, errors and calculating the SSE and Mean SE Null
# model
#
# In this code chunk you can write the functions that you need to use in order 
# to add the predictions and squared errors of the null model to the data.
# Calculate the SSE and the Mean SE from the new variables.
```

**ANS:** [WRITE YOUR ANSWER HERE]

18. What is the value of the Bayesian Information Criterion (BIC) for the Null model?

```{r bic-null-s2, echo = TRUE}
# Code 17: Calculating the BIC for the Null model
#
# In this code chunk you can write the functions that you need in order to 
# calculate the BIC associated to the Null model. Save it as a new variable and 
# report the value as text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

19. Using the values you obtained in question 16 obtain the predictions of the 
biological sex model, add a new column with the predictions, and calculate the squared error of each observation. Then obtain the SSE and the Mean SE associated with the biological sex model. Report the prediction, SSE and Mean SE of the Null model below.

```{r bio-m2-s2, echo = TRUE, fig.align = 'center'}
# Code 18: adding predictions, errors and calculating the SSE and Mean SE
# for the biological sex model
#
# In this code chunk you can write the functions that you need to use in order 
# to add the predictions and squared errors of the biological sex model to the
# data. Calculate the SSE and the Mean SE from the new variables.
```

20. What is the value of the Bayesian Information Criterion (BIC) for the 
biological sex model?

```{r bic-bio-s2, echo = TRUE}
# Code 19: Calculating the BIC for the biological sex model
#
# In this code chunk you can write the functions that you need in order to 
# calculate the BIC associated with the biological sex model. Save it as a new 
# variable and report the value as text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

21. Using the values you obtained in question 16, obtain the predictions of the 
word frequency model, add a new column with the predictions, and calculate the squared error of each observation. Then obtain the SSE and the Mean SE associated with the word frequency model. Report the prediction, SSE and Mean Se of the Null model below.

```{r freq-m2-s2, echo = TRUE, fig.align = 'center'}
# Code 18: adding predictions, errors and calculating the SSE and Mean SE
# for the word frequency model
#
# In this code chunk you can write the functions that you need to use in order 
# to add the predictions and squared errors of the word frequency model to the
# data. Calculate the SSE and the Mean SE from the new variables.
```

22. What is the value of the Bayesian Information Criterion (BIC) for the 
word frequency model?

```{r bic-bio-s2, echo = TRUE}
# Code 19: Calculating the BIC for the word frequency model
#
# In this code chunk you can write the functions that you need in order to 
# calculate the BIC associated with the word frequency model. Save it as a new  
# variable and report the value as text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

23. Using the values you obtained in question 16 obtain the predictions of the 
additive model, add a new column with the predictions, and calculate the squared
error of each observation. Then obtain the SSE and the Mean SE associated with 
the word frequency model. Report the prediction, SSE and Mean SE of the Null 
model below.

```{r freq-m2-s2, echo = TRUE, fig.align = 'center'}
# Code 18: adding predictions, errors and calculating the SSE and Mean SE
# for the additive model
#
# In this code chunk you can write the functions that you need to use in order 
# to add the predictions and squared errors of the additive model to the
# data. Calculate the SSE and the Mean SE from the new variables.
```

24. What is the value of the Bayesian Information Criterion (BIC) for the 
additive model?

```{r bic-bio-s2, echo = TRUE}
# Code 19: Calculating the BIC for the additive model
#
# In this code chunk you can write the functions that you need in order to 
# calculate the BIC associated with the additive model. Save it as a new  
# variable and report the value as text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

25. Which of the four models would you select and why?

**ANS:** [WRITE YOUR ANSWER HERE]


26. What does the Model that you selected tell us about our research question?

**ANS:** [WRITE YOUR ANSWER HERE]
