---
title: "homework 1"
author: "student name"
date: ""
output: pdf_document
---

## Section 1

**Background:** This first data set is from an experiment designed to see if 
drinking coffee makes people smarter. Participants were divided into two groups. 
All participants took an IQ test, but one group of participants drank water 5 
minutes before taking the IQ test (**group 1**). The other group drank coffee 
**group 2**.

```{r load_data, echo = TRUE}
# In this section you should write the code needed to load the data for
# section 2. The other calculations can be done in the following
# code chunks. Remember to delete the comments inside the code chuncks BEFORE 
# submitting your homework.
```

\small Questions 1 - 4 require you to use R code, questions 5 and 6 require you
to reflect on its outcome.

1. What is the average score on the *IQ* test for participants who did not drink
coffee?

```{r mean_group1, ehco = TRUE}
# In this code chunck you can write the function (or functions) that you need to 
# use in order to calculate the mean of the FIRST group, remember to save the 
# result in a new variable so that you can easily it in text. Remember to delete 
# the comments inside the code chuncks BEFORE submitting your homework.
```

**ANS:** 

1. What is the average score on the *IQ* test for participants who drinl coffee?

```{r mean_group2, ehco = TRUE}
# In this code chunck you can write the function (or functions) that you need to 
# use in order to calculate the mean of the SECOND group, remember to save the 
# result in a new variable so that you can easily it in text. Remember to delete 
# the comments inside the code chuncks BEFORE submitting your homework.
```

**ANS:** 

1. Which group had more variatiopn in their test scores on the *IQ* test, those 
who drink coffee or those who didn't? How do you know?

```{r variantion_group, echo = TRUE}
# In this code chink you can write the functions that you need to use in order 
# to calculate the variation of each group. Remember to assign each value to a 
# new variable so that you can use it in text.
```

**ANS:**


1. Create 2 histograms of the *IQ* test results of each group. One should show 
how participants who weren’t given coffee performed on the IQ test, and the 
other should show how participants who were given coffee performed. Describe 
what you see in the histograms in as few words as possible.

```{r histograms, echo = TRUE, fig.align = 'center'}
# Use this code chunk to make the histograms of the IQ test scores of each group
# it would be easier to compare results if you add transparency to the color! 
```

**ANS:** 

1. We take the mean IQ of participants who drank coffee so that we can get a single number that helps describe the performance of all those participants who drank coffee. We take the mean IQ of participants who didn’t drink coffee so that we can get a single number that helps describe the behavior of all those participants who didn’t drink coffee. Which do you think does a better job of describing the behavior of each respective group, the mean of those who drank coffee or the mean of those who didn’t? In other words, which mean seems more effective for describing the performance of participants in each condition? Why?

**ANS:**

1. If someone asked you, does drinking coffee make a difference for how people perform on IQ tests, what would you say and why?

**ANS:**



## Section 2

**Background:** This second data set comes from an experiment designed to see if 
a new form of training is effective in improving participant's ability to 
navigate a maze. Participants in the experimental condition first underwent the 
new training that might improve their performance. Participants in the control 
condition played an unrelated game that lasted approximately as long as the 
training. Participants in both conditions then completed the maze as quickly as 
possible. The data provides information about how quickly each participant 
completed the maze, in seconds.

```{r load_data2, echo = TRUE}
# In this section you should write the code needed to load the data for
# section 2. The other calculations can be done in the following
# code chunks. Remember to delete the comments inside the code chuncks BEFORE 
# submitting your homework.
```

\small Questions 7 - 10 require you to use R code, questions 11 and 13 require 
you to reflect on its outcome.

1. On average, how much did participants in the control condition improve by? Make sure your answer is clear, both in terms of units and $+/-$.


```{r diff_group, echo = TRUE}
# In this section you should write the code needed to calculate the average 
# improvement for both groups, remember to save the result in a new variable and 
# to use it in the text accessing the elements of the vector.
```

**ANS:**

1. On average, how much did participants in the experimental condition improve by? Again, make sure your answer is clear, both in terms of units and $+/-$.

**ANS:**

1. Which group had more variation in their improvement from one puzzle to the next, those who received the training or those who didn’t? How do you know?

```{r variantion_sec2, echo = TRUE}
# In this code chunk you can write the functions that you need to use in order 
# to calculate the variation of each group. Remember to assign each value to a 
# new variable so that you can use it in text.
```

**ANS:**

1. Create 2 boxplots. One should show how participants who weren’t given the 
training improved from the first puzzle to the second puzzle, and the other 
should show how participants who did get the training improved. These boxplots 
should be side by side for comparison purposes. Write a small comment on what 
you see.

```{r box-plots, echo = TRUE, fig.align = 'center'}
# Use this code chunk to make the boxplots of the average time to solve a maze 
# after training divided by group.
```

**ANS:**

1. Similar question 5 from Section 1, which mean do you think does a better job 
of describing participants behavior in each condition, the mean improvement of 
participants who did receive the training or the mean improvement of those who 
didn’t? Why?

**ANS:**

1. If someone asked you whether the training was effective in improving people’s performance on the logic puzzles, what would you say? Why?

**ANS:**

1. Considering the data you have looked at in this assignment, would you be more confident in saying that drinking coffee improves people’s performance on IQ tests or that the new training improves people’s performance on logic puzzles? Why?

**ANS:**