---
title: "Homework 2"
author: "Write your name here"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r load_data, echo = FALSE}
# Use this chunk to load whatever libraries you need. They should already have been
# installed for the first homework, but you need to load them using library()
# in order to use them. IMPORTANT: Remember to delete the comments inside the code
# chunks BEFORE submitting your homework.
```

## Section 1

**Background:** For this section, we will work with our coffee data set again. 
This data set comes from an experiment designed to test whether drinking coffee 
makes people smarter. Participants were divided into two groups. All 
participants took an IQ test, but participants in **group 1** 
drank water 5 minutes before the test, while participants in **group 2** drank 
coffee. 

We have two competing hypotheses (models). The first one is that 
drinking coffee doesn't make a difference in IQ test scores between people who 
drank water (group 1) and people who drank coffee (group 2). The second one 
is that people who drank coffee are smarter; in other words, the IQ 
test scores should be different between groups. For both hypotheses, we will 
assume that variances are equal across observations and groups.

```{r load_data, echo = TRUE}
# In this chunk you should write (a.k.a. copy from the slides) the code needed 
# to load the data for section 1.
```

1. Write the first model (**no differences between groups**) in mathematical
notation

**ANS:** [WRITE YOUR ANSWER HERE]

2. Make a density graph of model 1 (**no differences between groups**). Remember 
that at this point, the units (values of x) don't matter. We are only interested 
in what the model expects the data to look like. Assume that the model variance 
is equal to 3.

```{r density-m1-s1, echo = TRUE, fig.align = 'center'}
# In this code chunk you can write the code used to generate the density plot
# for model 1. Remember that the values on the x axis are not important. What we 
# are looking at is what the model expects the data to look like.
```

3. Calculate the predictions of the first model and add them as a new variable 
(column) to the data set (remember to give the variable a name!). What are the 
predictions of the first model?

```{r pred-m1-s1, echo = TRUE}
# In this code chunk you can write the functions that you need to use in order 
# to calculate the predictions of model one and add it to the data set as a new
# column. Report the value(s) as text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

4. Calculate the squared error (SE) of each observation for the first model and
add them to the data set as a new variable. What is the mean Sum of Squared Errors 
(m-SSE) of the "no group differences" model (model 1)?

```{r sse-m1-s1, echo = TRUE}
# In this code chunk you can write the functions that you need to use in order 
# to calculate the squared error of each observation and the functions you need 
# to calculate the sum of squared errors of the "no differences" model.
```

5. What is the value of the Bayesian Information Criterion for model 1?

```{r bic-m1-s1, echo = TRUE}
# In this code chunk you can write the functions that you need in order to 
# calculate the BIC associated with model 1. Save it as a new variable and 
# report the value as text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

6. Write the second model (**groups are different**) in mathematical notation.

**ANS:** [WRITE YOUR ANSWER HERE]

7. Make a density graph of model 2 (**groups are different**). Remember that at 
this point, the units (values of x) don't matter. We are only interested in what 
the model expects the data to look like. Assume that the model variance 
is equal to 3.

```{r density-m2-s1, echo = TRUE, fig.align = 'center'}
# In this code chunk you can write the code used to generate the density plot
# for model 2. Remember that the values on the x axis are not important. What we 
# are looking at is what the model expects the data to look like.
```

8. Calculate the predictions of the second model and add them as a new variable 
(column) to the data set (remember to give the variable a different name!). What 
are the predictions of the second model?

```{r pred-m2-s1, echo = TRUE}
# In this code chunk you can write the functions that you need to use in order 
# to calculate the predictions of model two and add it to the data set as a new
# column. Report the value(s) as text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

9. Calculate the squared error (SE) of each observation for the second model and
add them to the data set as a new variable. What is the mean Sum of Squared Errors 
(m-SSE) of this "group differences" model (model 2)?

```{r sse-m2-s1, echo = TRUE}
# In this code chunk you can write the functions that you need to use in order 
# to calculate the squared error of each observation and the functions you need 
# to calculate the mean sum of squared errors of the "group differences" model.
```

**ANS:** [WRITE YOUR ANSWER HERE]

10. What is the value of the Bayesian Information Criterion for model 2?

```{r bic-m2-s1, echo = TRUE}
# In this code chunk you can write the functions that you need in order to 
# calculate the BIC associated with model 2. Save it as a new variable and 
# report the value as text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

11. Which of the two models would you select and why?

**ANS:** [WRITE YOUR ANSWER HERE]



## Section 2

**Background:** For this section, we will work with our response time data set again.
This data set comes from an experiment designed to test whether a new training 
technique can improve participants' ability to navigate a maze. Participants 
first completed a maze after playing a game that lasted as long as the new 
training approach. Then they underwent the new training and completed a second 
maze of the same difficulty (we will assume that, if there is a difference in 
response times, it is only because of the training and not because of experience 
with maze solving). The data provides information about the time (in seconds) 
in which each participant completed each maze.

We have two competing models (hypotheses) about the performance of participants
on the maze before and after training. The first one (**no improvement**) states 
that the new training method does not make any difference. In other words, 
there will be no improvement in how quickly people solve the maze. 

The second model (**improvement**) states that the new training method is 
effective, and so the time people need to solve a maze should be shorter 
the second time (after participants have undergone the training).

```{r load_data2, echo = TRUE}
# In this section you should write the code needed to load the data for
# section 2. IMPORTANT: Remember to delete the comments inside the code 
# chunks BEFORE submitting your homework.
```

12. Write the first model (**no improvement**) in mathematical
notation.

**ANS:** [WRITE YOUR ANSWER HERE]

13. Make a density graph of model 1 (**no improvement**). To make the plot, 
assume that the model variance is equal to 2.

```{r density-m1-s2, echo = TRUE, fig.align = 'center'}
# In this code chunk you can write the code used to generate the density plot
# for model 1.
```

14. Write the predictions of the first model (**no improvement**).

```{r pred-m1-s2, echo = TRUE}
# In this code chunk you can write the functions that you need to use in order 
# to calculate the predictions of model 1 and add it to the data set as a new
# column. Report the value(s) as text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

15. Calculate the squared error (SE) of each observation for the first model and
add them to the data set as a new variable. What is the mean Sum of Squared Errors 
(m-SSE) of the "no improvement" model (model 1)?

```{r sse-m1-s2, echo = TRUE}
# In this code chunk you can write the functions that you need to use in order 
# to calculate the squared error of each observation and the functions you need 
# to calculate the sum of squared errors of the "no improvement" model.
```
**ANS:** [WRITE YOUR ANSWER HERE]

16. What is the value of the Bayesian Information Criterion for model 1?

```{r bic-m1-s2, echo = TRUE}
# In this code chunk you can write the functions that you need in order to 
# calculate the BIC associated to model 1. Save it as a new variable and 
# report the value as text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

17. Write the second model (**improvement**) in mathematical notation.

**ANS:** [WRITE YOUR ANSWER HERE]

18. Make a density graph of model 2 (**improvement**). Remember that at 
this point, the units (values of x) don't matter. We are only interested in what 
the model expects the data to look like.

```{r density-m2-s2, echo = TRUE, fig.align = 'center'}
# In this code chunk you can write the code used to generate the density plot
# for model 2. Remember that the values on the x axis are not important, what we 
# look at is what the model expects the data to look like.
```

19. Calculate the predictions of the second model and add them as a new variable 
(column) to the data set (remember to give the variable a different name!). What 
are the predictions of the second model?

```{r pred-m2-s2, echo = TRUE}
# In this code chunk you can write the functions that you need to use in order 
# to calculate the predictions of model 2 and add it to the data set as a new
# column. Report the value(s) as text below
```

**ANS:** [WRITE YOUR ANSWER HERE]

20. Calculate the squared error (SE) of each observation for the second model 
and add them to the data set as a new variable. What is the mean Sum of Squared 
Errors (m-SSE) of the "improvement" model (model 2)?

```{r sse-m2-s2, echo = TRUE}
# In this code chunk you can write the functions that you need to use in order 
# to calculate the squared error of each observation and the functions you need 
# to calculate the mean sum of squared errors of the "improvement" model.
```

**ANS:** [WRITE YOUR ANSWER HERE]

21. What is the value of the Bayesian Information Criterion for model 2?

```{r bic-m2-s2, echo = TRUE}
# In this code chunk you can write the functions that you need in order to 
# calculate the BIC associated with model 2. Save it as a new variable and 
# report the value as text below.
```

**ANS:** [WRITE YOUR ANSWER HERE]

22. Which of the two models would you select and why?

**ANS:** [WRITE YOUR ANSWER HERE]

