---
title: "homework 4"
author: "Manuel Villarreal"
date: "4/18/2022"
output: pdf_document
---

```{r load-library, echo = FALSE, message = FALSE, warning = FALSE}
# Code 0: loading tidyverse library

# This code chunk will load the tidyverse library but it will not 
# be displayed on the pdf.
library(tidyverse)
```

**Background:** For this section, we will work with the responses to a memory 
experiment. Researchers are interested in the effects of word frequency (how 
commonly a word is used in English) in two different tasks. A free recall task, 
where participants have to write down all the words they remember after studying
a list. A recognition task, where participants have to tell if they recognize a 
word presented to them as being part of a list that they studied. We have a 
2 by 2 factorial design. The first factor is the task (free recall vs 
recognition) and the second factor is word frequency (high vs low). The 
researchers recruited 100 participants for the study and assigned 25 to each 
combination of the factor levels (25 each combination) and the dependent 
variable was the number of words that where correctly recognized or correctly 
recalled.


```{r load_data, echo = TRUE}
# Code 1: Load the data (Hint: Go back to Lec-15 and use the link)
#
# In this section you should write the code needed to load the data for
# section 2. IMPORTANT: Remember to delete the comments inside the code 
# chunks BEFORE submitting your homework.
```

1. Obtain the predictions of the full model (cell means). What are the 
predictions of the **full** model for each combination of the levels of our 
factors *task* and *word frequency*?

```{r cell-means, echo = TRUE}
# Code 2: calculating the full model's predictions (cell means)
#
# In this code chunk you can write the functions that you need in order 
# to calculate the cell means (predictions of the full model).
```

**ANS:** [WRITE YOUR ANSWER HERE]

2. Using the cell means from (1) obtain the means for the levels of the *task* 
factor, consider *free recall* as the first level ($j=1$ if the observation
belongs to the free recall task). What is the mean of each *task*?

```{r task-mean, echo = TRUE}
# Code 3: average of the levels of the first factor.
#
# In this code chunk you can write the functions that you need to use in order 
# to calculate the average of each level of task (factor 1).
```

**ANS:** [WRITE YOUR ANSWER HERE]

3. Using the cell means from (1) obtain the means for the levels of the 
*word frequency* factor (factor 2), consider *low frequency* as the first level 
($k=1$ if the observation belongs to the free *low frequency* words). What is 
the mean of each *word frequency* list?

```{r wordfreq-mean, echo = TRUE}
# Code 4: average of the levels of the second factor
#
# In this code chunk you can write the functions that you need to use in order 
# to calculate the average of each level of word frequency (factor 2).
```

**ANS:** [WRITE YOUR ANSWER HERE]

4. Using the cell means from (1) obtain the grand mean. What is the value of the 
grand mean for this experiment and how can it be interpreted?

```{r grand-mean, echo = TRUE}
# Code 5: calculating the grand mean
#
# In this code chunk you can write the functions needed to calculate the grand 
# mean of the experiment.
```

5. Obtain the effects of the *task* factor ($\hat{\alpha}_j$). What are the 
values associated with each level of the task factor and how can we interpret 
them?

```{r me-task, echo = TRUE}
# Code 6: Calculating the effects of task factor
#
# In this code chunk you can write the functions that you need in order to 
# calculate the effects associated with the task factor.
```

**ANS:** [WRITE YOUR ANSWER HERE]

6. Obtain the effects of the *word frequency* factor ($\hat{\beta}_k$). What are 
the values associated with each level of the word frequency factor and how can 
we interpret them?

```{r me-wordfreq, echo = TRUE}
# Code 7: Calculating the effects of word frequency factor
#
# In this code chunk you can write the functions that you need in order to 
# calculate the effects associated with the word frequency factor.
```

**ANS:** [WRITE YOUR ANSWER HERE]

7. Add the predictions of the Null model to your data and calculate the squared 
error of each observation (your answer is just the code).

```{r pred-null, echo = TRUE}
# Code 8: Adding the predictions of the Null model to data and calculating
# the squared error by observation
#
# In this code chunk you can write the functions that you need in order to 
# add the prediction of the null model to the data and calculate the squared 
# error of each observation.
```

8. Calculate the predictions of the *task* main effects model and add them to 
the data. Use those predictions to calculate the squared error of each 
observation. What are the predictions of the *task* main effects model for each
combination of the levels of our factors?

```{r pred-me-task, echo = TRUE}
# Code 9: Calculating and adding the predictions of the task main effects model
# to the data and calculating the squared error by observation.
#
# In this code chunk you can write the functions that you need in order to 
# calculate and add your predictions of the task main effects model to the data. 
# Calculate the squared error of each observation and add it.
```

**ANS:** [WRITE YOUR ANSWER HERE]

9. Calculate the predictions of the *word frequency* main effects model and add 
them to the data. Use those predictions to calculate the squared error of each 
observation. What are the predictions of the *word frequency* main effects model 
for each combination of the levels of our factors?

```{r pred-me-wordfreq, echo = TRUE}
# Code 10: Calculating and adding the predictions of the word frequency main 
# effects model to the data and calculating the squared error by observation.
#
# In this code chunk you can write the functions that you need in order to 
# calculate and add your predictions of the word frequency main effects model 
# to the data. Calculate the squared error of each observation and add it.
```

**ANS:** [WRITE YOUR ANSWER HERE]

10. Calculate the predictions of the *additive* model and add them to the data.
Use those predictions to calculate the squared error of each observation. What 
are the predictions of the *additive* model for each combination of the levels 
of our factors?

```{r pred-additive, echo = TRUE}
# Code 11: Calculating and adding the predictions of the additive model to the 
# data and calculating the squared error by observation.
#
# In this code chunk you can write the functions that you need in order to 
# calculate and add your predictions of the additive model to the data. 
# Calculate the squared error of each observation and add it.
```

**ANS:** [WRITE YOUR ANSWER HERE]

11. Add the predictions of the *full* model to the data and use them to obtain 
the squared error by observation. What are the predictions of the *full* model
for each combination of the levels of our factors?

```{r pred-full, echo = TRUE}
# Code 12: Adding the predictions of the full model to the data and calculating 
# the squared error by observation.
#
# In this code chunk you can write the functions that you need in order to 
# add your predictions of the full model to the data and obtain the squared 
# error of each observation and add it.
```

**ANS:** [WRITE YOUR ANSWER HERE]
