---
title: "Homework 5 - Key"
author: "Manuel Villarreal"
date: "6/3/2022"
output: pdf_document
---

```{r load-library, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
```
```{r load_data, echo = TRUE, message=FALSE,warning=FALSE}
link <- "https://raw.githubusercontent.com/ManuelVU/psych-10c-data/main/homework5.csv"
grades <- read_csv(file = link)
```

1. **0.4 points**

```{r null, echo = TRUE}
betas_null <- lm(formula = grade ~ 1, data = grades)$coef
grades <- grades %>% 
  mutate("prediction_null" = betas_null,
         "error_null" = (grade - prediction_null)^2)
```

**ANS:** The prediction of the null model is equal to the estimate of $\beta_0$

| Model | Prediction | 
|-------|:----------:|
| Null  | `r round(betas_null,2)` |

2. **0.8 points**

```{r study-hours, echo = TRUE}
betas_sw <- lm(formula = grade ~ study_week, data = grades)$coef

grades <- grades %>% 
  mutate("prediction_sw" = betas_sw[1] + betas_sw[2] * study_week,
         "error_sw" = (grade - prediction_sw)^2)
```

**ANS:** The values of the parameters in the model are:

| Parameter | Estimate   |
|-----------|:----------:|
| $\beta_0$  | `r round(betas_sw[1],2)` |
| $\beta_1$  | `r round(betas_sw[2],2)` |


3. **0.8 points**

```{r classes-missed, echo = TRUE}
betas_cm <- lm(formula = grade ~ classes_missed, data = grades)$coef

grades <- grades %>% 
  mutate("prediction_cm" = betas_cm[1] + betas_cm[2] * classes_missed,
         "error_cm" = (grade - prediction_cm)^2)
```

**ANS:**  The estimated values of the parameters in the model are:

| Parameter | Estimate   |
|-----------|:----------:|
| $\beta_0$  | `r round(betas_cm[1],2)` |
| $\beta_1$  | `r round(betas_cm[2],2)` |

4. **0.8 points**

```{r grand-mean, echo = TRUE}
betas_swcm <- lm(formula = grade ~ study_week + classes_missed, data = grades)$coef

grades <- grades %>% 
  mutate("prediction_swcm" = betas_swcm[1] + betas_swcm[2] * study_week + 
           betas_swcm[3] * classes_missed,
         "error_swcm" = (grade - prediction_swcm)^2)
```

**ANS:** The estimated value of the parameters of this model are:

| Parameter | Estimate   | 
|-----------|:----------:|
| $\beta_0$  | `r round(betas_swcm[1],2)` |
| $\beta_1$  | `r round(betas_swcm[2],2)` |
| $\beta_2$  | `r round(betas_swcm[3],2)` |


5. **0.8 points**

```{r me-task, echo = TRUE}
grades <- grades %>% 
  mutate("classes_id" = case_when(classes_missed == 0 ~ 0,
                                  classes_missed > 0 ~ 1))
```

**ANS:** The proportion of students that missed at least one class was 
**`r round(mean(grades$classes_id),2)`**.

6. **0.8 points**
 
```{r me-wordfreq, echo = TRUE}
betas_swcid <- lm(formula = grade ~ study_week + classes_id, data = grades)$coef

grades <- grades %>% 
  mutate("prediction_swcid" = betas_swcid[1] + betas_swcid[2] * study_week + 
           betas_swcid[3] * classes_id,
         "error_swcid" = (grade - prediction_swcid)^2)
```

**ANS:** The estimated values of the parameters of the model are:

| Parameter | Estimate   | 
|-----------|:----------:|
| $\beta_0$  | `r round(betas_swcid[1],2)` |
| $\beta_1$  | `r round(betas_swcid[2],2)` |
| $\beta_2$  | `r round(betas_swcid[3],2)` |

7. **0.8 points**

```{r pred-null, echo = TRUE}
betas_int <- lm(formula = grade ~ study_week + classes_id + study_week * classes_id, data = grades)$coef

grades <- grades %>% 
  mutate("prediction_int" = betas_int[1] + betas_int[2] * study_week + 
           betas_int[3] * classes_id + 
           betas_int[4] * classes_id * study_week,
         "error_int" = (grade - prediction_int)^2)
```

**ANS:** The estimated values of the parameters of the model are:

| Parameter | Estimate   | 
|-----------|:----------:|
| $\beta_0$  | `r round(betas_int[1],2)` |
| $\beta_1$  | `r round(betas_int[2],2)` |
| $\beta_2$  | `r round(betas_int[3],2)` |
| $\beta_3$  | `r round(betas_int[4],2)` |

8. **0.8 points (code is the answer)**

```{r pred-me-task, echo = TRUE}
sse_null <- sum(grades$error_null)
sse_sw <- sum(grades$error_sw)
sse_cm <- sum(grades$error_cm)
sse_swcm <- sum(grades$error_swcm)
sse_swcid <- sum(grades$error_swcid)
sse_int <- sum(grades$error_int)
```

9. **0.8 points (code is the answer)**

```{r pred-me-wordfreq, echo = TRUE}
n_total <- nrow(grades)

mse_null <- 1/n_total * sse_null 
mse_sw <- 1/n_total * sse_sw 
mse_cm <- 1/n_total * sse_cm 
mse_swcm <- 1/n_total * sse_swcm 
mse_swcid <- 1/n_total * sse_swcid 
mse_int <- 1/n_total * sse_int 
```

10. **0.8 points**

```{r pred-additive, echo = TRUE}
bic_null <- n_total * log(mse_null) + 1 * log(n_total)
bic_sw <- n_total * log(mse_sw) + 2 * log(n_total) 
bic_cm <- n_total * log(mse_cm) + 2 * log(n_total)
bic_swcm <- n_total * log(mse_swcm) + 3 * log(n_total) 
bic_swcid <- n_total * log(mse_swcid) + 3 * log(n_total)
bic_int <- n_total * log(mse_int) + 4 * log(n_total)
```

**ANS:** The values of the BIC are the following (take 0.2 points if they 
miss the labels for at least one BIC value).

| Parameter | Estimate   | 
|-----------|:----------:|
| Null  | `r round(bic_null[1],2)` |
| Hours | `r round(bic_sw,2)` |
| Classes  | `r round(bic_cm,2)` |
| Hours + Classes  | `r round(bic_swcm,2)` |
| Hours + ID | `r round(bic_swcid,2)` |
| Interaction  | `r round(bic_int,2)` |


11. **0.8 points**

**ANS:** According to the BIC the best model is the interaction model, its
parameters can be interpreted as (order of the values depends on the order they 
assign in the lm function):

- $\beta_0$: expected grade of students that missed no classes and studied for 
0 hours each week.

- $\beta_1$: impact of an additional hour of study on the expected grade of 
students that missed 0 classes.

- $\beta_2$: difference on the expected grade of students who missed at least 
one class when they study for 0 hours a week in comparison to students that 
missed no classes.

- $\beta_3$ difference in the impact of an additional hour of study on the 
expected grade of students that missed at least one class in comparison to 
students that missed no classes.

12. **0.8 points**

```{r sse-5models, echo = TRUE}
r2_int <- (sse_null - sse_int) / sse_null
```

**ANS:** The proportion of variance accounted for by the model was **`r round(r2_int,2)`**.

13. **0.8 points**

**ANS:** Something along the lines of more hours of study are associated with 
better grades, however, the impact of each additional hour depends on whether 
students missed at least one class. Students that missed at least one class have 
a lower grade on average in comparison to students that missed no classes when 
they spend 0 hours a week studying.









